Typestate Inference in an Imperative Calculus

### 2013-01-02 formalism/paper.tex
progress and preservation theorem reviewed and fixed

\begin{abstract}
Current typestate research is primarily focused on studying
type systems that can support particular use cases for typestate;
the type systems are often complex, and programs in such languages 
require large type annotations to allow for full static type checking. Little
work has been done to determine whether inference of such annotations is 
feasible. We present a small imperative calculus which
models interactions with typestate-constrained objects and a type inference
technique which can derive the minimal state machine that governs interactions
with the observed objects, and therefore eliminate the need for
explicit type annotations.
\end{abstract}

@@

An object may have a very simple 
protocol, where the set of messages (or equivalently, method calls) it is 
willing to accept is constant throughout its lifecycle --- this is what the 
type system of programming languages such as Java can express. This is
insufficient to correctly model the object's behaviour in many cases, as
the set of messages the object can correctly respond to is dependent upon an
abstraction of its private state. State change is typically synchronous, in 
response to message receipt, and safety requires that 
any message sent to an object is within the set which can be received for
the currently known state.

Describing the allowed sequence of messages which a class of objects can
receive, their typestate, produces significantly more complex types than
in an object oriented language without typestate constraints. Functions which
take objects as parameters must describe the type of object they accept,
in which state, and what state transformation will occur (which may not
be deterministic). This type annotation burden is particularly pronounced
in a language with function literals, which must be succinct
in order to be readable.

Additionally, static checking of typestate requires
that the current state of an object always be known at a point of invocation.
This is only possible when an alias control strategy is employed, such as linear types, 
but such types prohibit programming in a style that most programmers are 
familiar with. Attempting to relax linear typing through the use of fractional 
permissions as in Plural and Plaid \cite{Aldrich2009,Bierhoff2009}
adds a significant amount of complexity to the type system. This results in 
large type annotations within programs to express both typestate and aliasing 
constraints when modular static type checking is desired.
If it were possible to eliminate the need for such type annotations, code
could be more concise and type annotations would only be necessary for
documentation purposes, and could be checked that they are simply more specific
versions of the principal typing produced by the type inference algorithm.

@@

A perhaps more intuitive basis for subtyping
uses the notion of {\it trace inclusion}. The set of interaction traces 
$\Tr(O@S)$ of a type is the smallest set such that:

@@

Either view of subtyping is also easy to relate to the Liskov substitution 
principle \cite{Liskov1987} --- a type will accept every sequence of
method calls that a supertype would, therefore it is a {\it behavioural}
subtype.

@@

TS is a small imperative calculus which models interactions with 
typestate-constrained objects, supports function literals without 
implicit capture, and employs structural subtyping for function and object
types.
The grammar of the language is shown 
in \figref{fig:grammar}, with operational semantics in \figref{fig:opsemantics} 
and typing rules in \figref{fig:typerules}.
The language is designed primarily
to study typestate inference in imperative code, rather than to 
be practical, as it uses syntactic control of interference \cite{Reynolds1978} 
to avoid the issue of alias control entirely - there is no means within the
language to produce an alias to an object.

@@

\begin{itemize}
\item {\tt unit}, the sole value of the {\tt Unit} type, which is used as
the result type for terms which have no meaningful result.
\item Function literals, which do not allow for implicit capture
of variables into their scopes --- everything must be passed as a parameter.
Tracking effects on implicitly captured references is not very well
studied in the current literature outside of strict linear types where
function literals that capture linear references must themselves be linear.
Parameters are passed by reference. 
Requiring that all references be explicitly passed to the function allows for
effects to be explicitly declared. The notation $T \gg T'$ is used to
describe an {\it effect type}, which declares a required type for the parameter 
on the left, and the new type of the parameter after application of the function 
on the right. Not all effect types are legal, as will be discussed later.
\item Object literals. These are composed of a state machine description
and a current state. The state machine description contains a set of
distinct state labels $\overline{S_i}$, within which is a set of methods $\overline{m_{ij}}$ 
which can be called in that state. The language does
not attempt to fully model the behaviour of objects, only interactions with
them. As such, it is sufficient for each method to be defined as a pair 
consisting of the return value of the method, and the next state of the object 
if the method is invoked. Non-determinism in the production of a return value 
could be simulated by specifying a set of possible return values and allowing 
the operational semantics to select one of these randomly.
All values would have to have the same type, or types for which a least upper 
bound exists. Such non-determinism would have no effect on the type system,
which is the main focus of this paper.
\end{itemize}

@@

\begin{itemize}
\item A conditional of form $\ift{t_c}{t_t}{t_f}$, which evaluates
$t_c$ first to a boolean result which decides which of $t_t$ 
(the \"true\" branch) or $t_f$ (the \"false\" branch) is
evaluated.
\item Function calls of form $x(\overline{x_i})$, which extract the function
literal stored in $x$ and apply it to the set of parameters $\overline{x_i}$,
which must match the declared parameter types. Each $x_i$ must be a distinct
variable, and is passed by reference, allowing the type of $x_i$ to be changed
as a result of the function application.
\item Method invocations on objects of form $x.m$. A method call is only
valid when the method is available in the current state of the object stored
in $x$. After a valid invocation, the state of the object will have changed
as specified by the object's state machine.
\item A let-bind construct $\lett{x}{t}{t'}$, which creates a new variable
$x$ within the scope of the body term $t'$ which is a reference to the value
derived from the term $t$. Rebinding a variable name is not allowed.
\item Sequencing of form $\tseq{t}{t'}$, which evaluates terms in 
left-to-right order.
\end{itemize}

@@

The example in \figref{fig:exampleprog} shows a short TS0 program
that TODO.

\begin{figure}
\begin{lstlisting}
TODO
\end{lstlisting}
\caption{\label{fig:exampleprog} An example program in TS0}
\end{figure}

@@

As there is no way for a reference to be duplicated, the mapping
from variable names to store locations is injective and therefore we have
no need for explicit ``location values'' as in the typical treatment of 
languages with references (such as in \cite{Pierce2002}). There is no need
for a separate store typing, as a well-typed store can be defined directly
in relation to a context as in \figref{fig:storetype}. A well-typed store
may contain mappings for variables which are not in the context as shown in
rule \textsc{st\_extra}.

@@

\subsection{Type rules and soundness}

@@

@@

Corresponding to the values of the language, there are four main kinds
in the type system: Unit, Bool, function types and object types. 
Two modifications are made to the definition of object protocols described
in \secref{sec:tsmodel}: methods have return types, and an object type
is the combination of an object protocol and a {\it set} of states.
An object protocol $O$ is therefore formally defined as the tuple 
$O = \langle M, \Sigma, \Delta \rangle$
where the transition relation $\Delta \subseteq \Sigma \times M \times T \times \Sigma$, 
where $T$ is the set of all types in the language.

While an object value in TS is still deterministic and has a single state at any 
given
point in time, we may not be able to determine in the type system which state 
exactly it is in --- conditional evaluation readily produces this situation.
With an if-then-else statement where the branches
interact with an object differently, it is common for the state of the object
at the end of each branch to be different. As
knowing precisely which branch will be taken is rarely possible in a
decidable type system, we must either insist that the state of the object
be the same upon exit from both branches, or devise a means to cope with
the object being in one of multiple potential states. We choose the latter
approach, as it is both more flexible and allows for intuitive principal
types to be chosen, as will be discussed later.

@@

An object type in TS, written $O@\overline{S}$, is equivalent to
$\bigsqcup_{S \in \overline{S}} O@S$, where $T \sqcup T'$ is the least upper
bound (or {\it join}) of $T$ and $T'$. Deriving a least upper bound for
two arbitrary object types (with a single current state) 
involves constructing the intersection of the state machines for each object 
type, using the current state of each object type as the initial state.
Similarly, deriving a greatest lower bound is based upon the union of
the two state machines in the same fashion. Join is always defined for
object types, while meet is not --- where the return types of methods are
incompatible (i.e. no join/meet exists respectively), a join operation will 
discard the method (it is unsafe to
call it) while a meet operation would require that the method be callable,
however no valid return type can be assigned to the method. Therefore
the kind of object types is a join semi-lattice, with the object type 
$\{ S_1 \{\} \}@S_1$ serving as the least element.

The intuition here is that the only sequences of method calls which can be 
safely invoked on an object of type $O@\overline{S}$ are those which could be 
safely invoked on each individual $O@S$, where $S \in \overline{S}$, and that
a least upper bound of the return types exists.

In general, the set of
types in TS is a join semi-lattice, with $\top$ as the greatest element.

@@

Given the formal definition of an object type $O@\overline{S}$, to save
referring to the $\Sigma$, $M$ and $\Delta$ that $O$ is composed of we shall
use the following notational conventions:

\begin{itemize}
\item $S \in O$ is the {\it state existence predicate} asserting 
$S$ is a state of $O$. 
Formally, this is equivalent to $S \in \Sigma$.
\item $m : T \Rightarrow S' \in O@S$ is the {\it method existence predicate} 
asserting the existence
of method $m$ with return type $T$ in state $S$ which triggers a 
transition to state $S'$ when called (where both $S \in O$ and $S' \in O$). 
Formally, this is equivalent to $(S,m,T,S') \in \Delta$.
\item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
of the method existence predicate to state sets, meaning that for all 
$S \in \overline{S}$ there exists a state and type 
$T'$ where $(S,m,T',S') \in \Delta$ and that
$T = \bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$ exists
and that $\overline{S'} = \{ S' | (S,m,T',S') \in \Delta \}$.
\item $O@\overline{S} \precsim O_2@\overline{S_2}$ is the extension of the
simulation relation to support state sets, and can be interpreted as
$\bigsqcup_{S \in \overline{S}}O@S \precsim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
\end{itemize}

@@

An ``effect type'' $T \gg T'$ describes how the type of a variable changes
in response to function application. In the TS language, the only type
changes that are those which change the state of an object instance; the
values of the other types are immutable. If a boolean is passed as a parameter,
the effect type will always be $\boolt \gg \boolt$, similarly for
$\unitt$ and $\funt{T}{U}{V}$. An effect type of $\boolt \gg \unitt$ is not
possible. In general, effect types are kind-preserving.

\begin{figure}
\includegraphics[width=0.49\textwidth]{sm1.pdf}
\caption{\label{fig:sm1} Operational semantics}
\end{figure}

In the case of an object parameter, not all effect types are possible
either. Consider
the object protocol in \figref{fig:sm1}, which we will refer to as $O_1$.
An effect type of $O_1@B \gg O_1@A$ is impossible for this protocol, as there
is no sequence of method calls which would result in a transition from state
$B$ to state $A$. An effect type of $O_1@B \gg O_1@\{B,D\}$ is possible, 
as there is at least one sequence of method calls on an object in
state $B$ to both state $B$ and $D$.
An effect type $O_1@{B,D} \gg O_1@{B,C}$ which has a state set on
the left hand side is possible as there exists a sequence of method calls 
that would take an object in state $B$ to $C$ that would also
take an object in state $D$ to $B$ (specifically, a single method
call to $b$).

In the general case, an effect type concerning an object protocol $O$
of form $O@\overline{S} \gg O@\overline{S_2}$ is
legal if:

@@

Checking the legality of an effect type is straightforward: by computing the
intersection of all the states, we can associate with each state in the 
intersection a set of states in the original machine
it was derived from (as shown in \figref{fig:sm1}. 
Should an associated state set contain an end state,
if all other states in the set are also end states then this represents a
``candidate'' end state within the intersection machine. The union of the state 
sets associated with all candidate end states must be equal to the
desired end state set for the effect type to be legal. This would correctly rule 
out the effect type $O@\{B,D\} \gg O@\{B\}$ (there is no sequence of method
calls that would take an instance in either state $B$ or $D$ and result in
the object exclusively being in state $B$) while accepting effect types
$O@\{B,D\} \gg O@\{B,D\}$ (corresponding to no operations occurring) and
$O@\{B,D\} \gg O@\{B,C,D\}$ (corresponding to either no operations occuring
or a single call to $b$).

\subsection{Principal types}

A principal type is one which makes the smallest commitment about 
the structure of a value based on its usage; due to the presence of subtyping 
in TS, a principal type is an upper bound on the type of the value.

\begin{figure}
\includegraphics[width=0.49\textwidth]{principal.pdf}
\caption{\label{fig:principalex}}
\end{figure}

Following from this general definition, a principal object type in TS 
is an object protocol and initial state set which permits all sequences of 
method calls that could be generated by an associated term $t$, and no more. 
In other words,
the language accepted by the state machine is the smallest possible, while
the return types of methods are as general as possible. Consider the example
function $f$ shown in \figref{lst:principalex}. It is clear from the usage
of $y$ that it must be of type $\boolt$ as demanded by the type judgement
of an if-then-else statement, while $x$ is an object of some sort.
We could ascribe any of the object types $O_1@A$, $O_2@A$ and $O_3@A$ to the parameter 
$x$, and after evaluation the object types would be $O_1@A$, $O_2@\{E,H\}$
and $O_2@F$ respectively.
$O_1@A$ is not principal as it will admit many more sequences of method calls 
than strictly required by the body of function $f$, while $O_2@A$ and $O_3@A$ 
are principal.

$O_2$ contains more states than are strictly necessary, i.e. it
is not a {\tt minimal} state machine, while $O_3$ is minimal.
We do not require that a principal type be minimal. If necessary, any
principal state machine we produce can be minimised, so this can be done
on demand where there is a specific need for it. One such need may be when
presenting the type to a user, as a minimal state machine may be easier to 
understand. Regardless, the state machines $O_2$ and $O_3$ can be shown to be 
isomorphic in the language that they accept (starting at state $A$), and that 
language is minimal, so we assert that either are acceptable principal types.

Similarly, it is possible for a principal effect type to have more than
one state in the state set of the input (left) object type. It is
always possible to restructure such an object type to have a single initial
state. It is not, however, possible to have both a single input state and
single output state for a state machine accepting an arbitrary language --- 
consider the language accepted by the regular expression $m^* | n^*$. It is 
necessary for a state machine which accepts this language, and only this 
language, to
have at least three exit states. In general, it is more intuitive to have
a single input state in the effect type but this is not required for a
type to be considered principal.

A principal type we can therefore ascribe to the function $f$
is $\forall T_1 \ldots T_6 . (O_3@A \gg O_3@F, \boolt) \rightarrow T_6$, 
or equivalently 
$\forall T_1 \ldots T_6 . (O_2@A \gg O_2@\{E,H\}, \boolt) \rightarrow T_6$.

There are occasional advantages to not always minimising principal state
machines. The additional structure of $O_2$ retains additional information
in the effect type which can be helpful when interpreting an effect type
when a subtype is passed to the function, as will be described in the next
section.

\subsection{Interpreting effect types}

The typing judgement for a function call permits a passed parameter to
be a subtype of the input portion of the associated effect type.
Determining a precise upper bound for the type of the parameter after the
function call, in particular where the parameter is an object, 
requires some thought.

Consider a function call $f(x)$ where
$f : (O_2@\overline{S_2} \gg O@\overline{S_4}) \rightarrow T$ and
$x : O_1@\overline{S} <: O_2@\overline{S_2}$, where $O_1$ and $O_2$
are the state machines from \figref{fig:principalex}, 
$\overline{S} = \{ A \}$, $S_2 = \{ A \}$ and $S_4 = \{E, H\}$. 
Based on the effect type,
we can assert that the function will have interacted with the $x$ using
one of the possible sequences that would be accepted by $O_2@\overline{S_2}$,
ending with the object in the state set $\overline{S_4}$.

A straightforward but questionable choice would 
be to replace the type of the variable
with that of the output portion of the effect type after the call. However,
this will often result in an unacceptable loss of type information,
reducing the number of valid programs which can be typed within the system.

It is clear that as $\ot{O}{S} \precsim \ot{O_2}{S_2}$ there must exist some
$\overline{S_3}$ such that $\ot{O}{S_3} \precsim \ot{O_2}{S_4}$.
One possible solution would be to choose
$\overline{S_3} = \{ S | O@S \precsim \ot{O_2}{S_4} \}$. This is not specific
enough however --- given the object protocols from \figref{fig:principalex},
$\overline{S_3} = \{ A, B, C, D \}$ as there are no possible method calls
on states $E$ or $H$ of $O_2$.

It is clear that our choice of $S_3$ must relate the set of possible method
call sequences between $\ot{O_2}{S_2}$ and $\ot{O_2}{S_4}$. For this we
define a relation between pairs of object types
$(O@\overline{S}, O_2@\overline{S_2} \lesssim 
(O@\overline{S'}, O_2@\overline{S_2'})$, which we call the
{\it simulation predecessor relation}. The definition of this relation
is given in \figref{fig:simreach}. Essentially, if $O@S \precsim O_2@S_2$ and there
exists a sequence of method calls that would transition an object value of type
$O_2@\overline{S_2}$ to $O_2@\overline{S_4}$, then this sequence of method calls
would transition an object in state $O@\overline{S}$ to $O@\overline{S_3}$
such that $O@\overline{S_3} \precsim O@\overline{S_4}$. This is abbreviated to
$(O@\overline{S}, O_2@\overline{S_2}) \lesssim (O@\overline{S_3}, O@\overline{S_4})$.

Using this relation, we can choose $\overline{S_3} = 
\bigcup \{ \overline{S'} | 
  (\ot{O}{S} , \ot{O_2}{S_2}) \lesssim (\ot{O}{S'}, \ot{O_2}{S_4})
\}$. The solution using the object protocols from \figref{fig:principalex}
will therefore be $\overline{S_3} = \{ A \}$.

@@

\subsection{Soundness}

The soundness of the type system is stated formally in the following
theorem:

\begin{thm}[Progress and Preservation]
Given a term $t$ such that $\typerule{\Gamma}{t}{T}{\Gamma'}$
with a store $\mu$ such that $\Gamma \vdash \mu$, either $t$ is a value or 
there exist $t'$ and $\mu'$ such that
$\oprule{t}{\mu}{t'}{\mu'}$, with some $\Gamma''$ such that
$\Gamma'' \vdash \mu'$ and $\typerule{\Gamma''}{t'}{T}{\Gamma'}$.
\end{thm}

\begin{proof}
in appendix \appref{app:soundness}.
\end{proof}

>>>>

\begin{abstract}
Current typestate research is primarily focused on studying
type systems that can support particular use cases;
the type systems are often complex, and programs in such languages 
require large type annotations to allow for full static type checking. Little
work has been done to determine whether inference of such annotations is 
feasible. We present a small imperative calculus which
models interactions with typestate-constrained objects and a type inference
technique which can derive principal typings for terms with some
restrictions, and therefore eliminate the need for
explicit type annotations in most cases.
\end{abstract}

@@

An object may have a very simple 
protocol, where the set of messages (or equivalently, method calls) it is 
willing to accept is constant throughout its lifecycle --- this is what the 
type system of programming languages such as Java can express. This is
insufficient to correctly model the object's behaviour in many cases, as
the set of messages the object can correctly respond to is dependent upon
its private state. State change is typically synchronous, in 
response to message receipt, and safety requires that 
any message sent to an object is within the set which can be received for
the current state. The private state can often be abstracted
such that restrictions on sequences of messages can be expressed as a
deterministic finite state machine.

Describing the allowed sequence of messages which a class of objects can
receive, their typestate, produces significantly more complex types than
in an object oriented language without typestate constraints. Functions which
take objects as parameters must describe the type of object they accept,
in which state, and what state transformation will occur (which may not
be deterministic). This type annotation burden is particularly heavy
in a language with function literals, which must be succinct
in order to be practical.

Static checking of typestate requires
that the current state of an object always be known at a point of invocation.
This is only possible when an alias control strategy is employed, such as linear
types, but such types prohibit programming in a style that most programmers are 
familiar with. Attempting to relax linear typing through the use of fractional 
permissions as in Plural and Plaid \cite{Aldrich2009,Bierhoff2009}
adds further complexity to the type system. This results in 
large type annotations within programs to express both typestate and aliasing 
constraints when modular static type checking is desired.
If it were possible to eliminate the need for such type annotations in
most situations, code could be more concise and readable, with annotations
reserved for situations where inference is not possible or are required
for formal documentation purposes.

@@

A perhaps more intuitive basis for subtyping is {\it trace inclusion}. The set 
of interaction traces $\Tr(O@S)$ of a type is the smallest set such that:

@@

Either view of subtyping is comparable to the Liskov substitution 
principle \cite{Liskov1987} --- a type must accept every sequence of
method calls that a supertype would, therefore it is a {\it behavioural}
subtype.

@@

TS is a small imperative calculus which models interactions with 
typestate-constrained objects, supports function literals without 
implicit capture, and employs structural subtyping for function and object
types.
The grammar of the language is shown 
in \figref{fig:grammar}, with operational semantics in \figref{fig:opsemantics} 
and typing rules in \figref{fig:typerules}.
The language is designed primarily
to study typestate inference in imperative code, rather than to 
be practical, general purpose programming language. The language uses syntactic 
control of interference \cite{Reynolds1978} to avoid the issue of alias control 
entirely - there is no means within the language to produce an alias to an 
object. This focuses attention on the fundamentals of typestate inference.

@@

\begin{itemize}
\item {\tt unit}, the sole value of the {\tt Unit} type, which is used as
the result type for terms which have no meaningful result.

\item {\tt true} and {\tt false}, of type {\tt Bool}, which are used for
decisions in conditional evaluation. The language contains no operations
for manipulating booleans, to keep the calculus small, however
such operations would be straightforward to add.

\item Function literals of form $\funv{x_i}{T_i}{T_i'}{t}$, where
$\overline{x_i}$ is the set of parameters to the function while
$t$ is the body term. Parameters are passed by reference.
Function literals do not allow for implicit capture
of variables into their scopes --- everything must be passed as a parameter. 
Tracking effects on implicitly captured references is not well understood
in the current literature outside of strict linear types where
function literals that capture linear references must themselves be linear.
Requiring that all references be explicitly passed to the function allows for
effects to be explicitly declared. The notation $T \gg T'$ is used to
describe an {\it effect type}, which declares a required type for the parameter 
on the left, and the new type of the parameter after application of the function 
on the right. Not all effect types are legal, as will be discussed later.

\item Object literals of form 
$O@S$, where $O = \left[ \overline{ S_i \{ m_{ij} = (v_{ij}, S_{ij}) \} } \right]$. 
$O$ is the object protocol, and $S$ is the current state
(where $S \in \overline{S_i}$). The object protocol contains a set of
distinct state labels $\overline{S_i}$, within which is a set of methods 
$\overline{m_{ij}}$ 
which can be called in that state. The language does
not attempt to fully model the behaviour of objects, only interactions with
them. As such, it is sufficient for each method to be defined as a pair 
consisting of the return value of the method, and the next state of the object 
if the method is invoked. Non-determinism in the production of a return value 
could be simulated by specifying a set of possible return values and allowing 
the operational semantics to select one of these randomly.
All values would have to have the same type, or types for which a least upper 
bound exists. Such non-determinism would have no effect on the type system,
which is the main focus of this paper.
\end{itemize}

@@

\begin{itemize}
\item A conditional of form $\ift{t_c}{t_t}{t_f}$, which evaluates
$t_c$ first to a boolean result which decides which of $t_t$ 
(the \"true\" branch) or $t_f$ (the \"false\" branch) is
evaluated.
\item Function calls of form $x(\overline{x_i})$, which extract the function
literal stored in $x$ and apply it to the set of parameters $\overline{x_i}$,
which must match the declared parameter types. Each $x_i$ must be a distinct
variable, and is passed by reference, allowing the type of $x_i$ to be changed
as a result of the function application.
\item Method invocations on objects of form $x.m$. A method call is only
valid when the method is available in the current state of the object stored
in $x$. After a valid invocation, the state of the object will have changed
as specified by the object's protocol.
\item A let-bind construct $\lett{x}{t}{t'}$, which creates a new variable
$x$ within the scope of the body term $t'$ which is a reference to the value
derived from the term $t$. Rebinding a variable name is not allowed.
\item Sequencing of form $\tseq{t}{t'}$, which evaluates terms in 
left-to-right order.
\end{itemize}

@@

The example in \figref{fig:exampleprog} shows a short TS0 program
that emulates the usage of an iterator. The object {\tt it}
provides an iterator-like contract, starting in the state {\tt CHECK} where
only the method {\tt hasNext} can be called, which returns a boolean value.
This method changes the state of the object to {\tt READ}, where both
{\tt hasNext} and {\tt next} can be called. Calling {\tt next} changes the
state of the object back to {\tt CHECK}. While the language does not
support {\it conditional transitions} that would be necessary to express
the restriction that {\tt next} should only be callable after {\tt hasNext}
returns {\tt true}, it does still provide more protection than a language
without typestate would.

\begin{figure}
\begin{lstlisting}
let it = [ CHECK { hasNext = (true,READ) } 
          READ  { hasNext = (true,READ);
                  next = (unit,CHECK) }
        ]@CHECK in
let console = [S { print = (unit,S) }]@S in
let f = \(x).(if x.hasNext 
              then (x.next ; true) 
              else false) in
if f(it) then console.print else unit
\end{lstlisting}
\caption{\label{fig:exampleprog} An example program in TS0}
\end{figure}

The object {\tt console} is effectively stateless, allowing the method
{\tt print} to be called without restriction.

The function {\tt f} will call {\tt hasNext}, and if it returns true then
will call {\tt next}. It returns a boolean value indicating whether
{\tt next} was called.

Finally, the object {\tt it} is passed to {\tt f} and {\tt print} will
be called on {\tt console} if {\tt f} returns true. Given the fixed return
values for the methods on {\tt it}, {\tt print} will always be called.

The most interesting part of the program is the function literal assigned
to {\tt f}. What is the effect type for the parameter {\tt x}? 
This is the only piece of type information that cannot be determined
easily, and is key to determining whether {\tt f} would use {\tt it} in
a safe manner.

@@

As there is no way for a reference to be duplicated, the mapping
from variable names to store locations is injective and therefore we have
no need for explicit ``location values'' as in the typical treatment of 
languages with references (such as in \cite{Pierce2002}). There is no need
for a separate store typing, as a well-typed store can be defined directly
in relation to a context as in \figref{fig:storetype}. A well-typed store
may contain mappings for variables which are not in the context as shown in
rule \textsc{st\_extra}. Such mappings are candidates for garbage collection,
or an explicit deallocation instruction could be injected after the
evaluation of the body of a let-binding.

@@

\subsection{Type system}

@@

The type system of TS is sound (it has both progress and type preservation
properties), which is proven in \thmref{thm:soundness} in 
\appref{app:soundness}.

@@

There are four kinds in the type system: Unit, Bool, function types and 
object types.
Two modifications are made to the definition of object protocols described
in \secref{sec:tsmodel}: methods have return types, and an object type
is the combination of an object protocol and a {\it set} of states.
An object protocol $O$ is therefore formally defined as the tuple 
$O = \langle M, \Sigma, \Delta \rangle$
where the transition relation $\Delta \subseteq \Sigma \times M \times T \times \Sigma$, 
where $T$ is the set of all types in the language.

While an object value in TS is still deterministic and has a single state at any 
given
point in time, we may not be able to determine in the type system which state 
exactly it is in --- conditional evaluation readily produces this situation.
With an if-then-else expression where the branches
interact with an object differently, it is common for the state of the object
at the end of each branch to be different. As
knowing precisely which branch will be taken statically is rarely possible, 
we must either insist that the state of the object
be the same upon exit from both branches, or devise a means to cope with
the object being in one of multiple potential states. We choose the latter
approach, as it is both more flexible and allows for intuitive principal
types to be chosen, as will be discussed later.

@@

An object type in TS, written $O@\overline{S}$, is equivalent to
$\bigsqcup_{S \in \overline{S}} O@S$, where $T \sqcup T'$ is the least upper
bound (or {\it join}) of $T$ and $T'$. Deriving a least upper bound for
two arbitrary object types (with a single current state) 
involves constructing the intersection of the state machines for each object 
type, using the current state of each object type as the initial state.
Similarly, deriving a greatest lower bound is based upon the union of
the two state machines in the same fashion. Join is always defined for
object types, while meet is not. Consider the two object types
$\{ S \{ m : \unitt \Rightarrow S \} \}@S$ and
$\{ S \{ m : \boolt \Rightarrow S \} \}@S$. The meet of these two types
requires that method $m$ is callable, but $\unitt \sqcap \boolt$ is undefined
and therefore no return type can be assigned to the method $m$.

The intuition for state sets is that the only sequences of method calls which 
can be safely invoked on an object of type $O@\overline{S}$ are those which 
could be safely invoked on each individual $O@S$, where $S \in \overline{S}$.

In general, the set of
types in TS is a join semi-lattice, with $\top$ as the greatest element.
The greatest element within object types exclusively is $\{ S \{\} \}@S$,
that is, an object which allows no method calls.

@@

Given an object type $O@\overline{S}$, to save
referring to the $\Sigma$, $M$ and $\Delta$ that formally define $O$
the following notational conventions shall be used:

\begin{itemize}
\item $S \in O$ is the {\it state existence predicate} asserting 
$S$ is a state of $O$. 
Formally, this is equivalent to $S \in \Sigma$.
\item $m : T \Rightarrow S' \in O@S$ is the {\it method existence predicate} 
asserting the existence
of method $m$ with return type $T$ in state $S$, which triggers a 
transition to state $S'$ when called (where both $S \in O$ and $S' \in O$). 
Formally, this is equivalent to $(S,m,T,S') \in \Delta$.
\item $m : T \Rightarrow \overline{S'} \in O@\overline{S}$ is the extension
of the method existence predicate to state sets, meaning that for all 
$S \in \overline{S}$ there exists a state $S'$ and type 
$T'$ where $(S,m,T',S') \in \Delta$.
$T$ is defined as 
$\bigsqcup_{S \in \overline{S}} \{ T' | (S,m,T',S') \in \Delta \}$
and $\overline{S'}$ is defined as $\{ S' | (S,m,T',S') \in \Delta \}$.
\item $O@\overline{S} \precsim O_2@\overline{S_2}$ is the extension of the
simulation relation to support state sets, and can be interpreted as
$\bigsqcup_{S \in \overline{S}}O@S \precsim \bigsqcup_{S_2 \in \overline{S_2}}O_2@S_2$.
\end{itemize}

\subsection{Object type isomorphism}

As object types in TS are based upon deterministic finite state machines,
there are often multiple ways to represent the same type. Object types
$O_1@S$ and $O_2@S$ are considered {\it isomorphic} 
(up to $\alpha$-conversion of state labels)
if they are both simulations of each other:
$O_1@\overline{S_1} \precsim O_2@S_2 \wedge O_2@\overline{S_2} \precsim O_1@\overline{S_1}$. 
The
{\it canonical} form amongst a set of isomorphic object types has a
{\it minimal} finite state machine.

@@

An ``effect type'' $T \gg T'$ describes how the type of a variable changes
in response to function application. In the TS language, as the values of
all types with the exception of objects are immutable, effects are
either identities ($T \gg T$) or object state changes 
($O@\overline{S_1} \gg O@\overline{S_2}$). As references cannot
be overwritten, we are guaranteed that the object instance cannot be replaced.
Therefore an effect type $O_1@\overline{S_1} \gg O_2@\overline{S_2}$ can
only be valid if $O_1 = O_2$.

\begin{figure}
\includegraphics[width=0.49\textwidth]{sm1.pdf}
\caption{\label{fig:sm1} A simple object protocol}
\end{figure}

State set transformations are restricted for semantic reasons also. Consider
the object protocol in \figref{fig:sm1}, which we will refer to as $O_1$.
An effect type of $O_1@B \gg O_1@A$ is impossible for this protocol, as there
is no sequence of method calls which would result in a transition from state
$B$ to state $A$. An effect type of $O_1@B \gg O_1@\{B,D\}$ is possible, 
as there is at least one sequence of method calls on an object in
state $B$ to both state $B$ and $D$.
An effect type $O_1@{B,D} \gg O_1@{B,C}$ which has more than one input state
is possible as there exists a sequence of method calls 
that would take an object in state $B$ to $C$ that would also
take an object in state $D$ to $B$ (specifically, a single method
call to $b$, which we can treat as having return type $\top = \boolt \sqcup \unitt$).

In the general case, an effect type concerning an object protocol $O$
of form $O@\overline{S} \gg O@\overline{S_2}$ is
valid if:

@@

Checking the validity of an effect type is straightforward: given an
effect type $O@\overline{S_1} \gg O@\overline{S_2}$, we first compute
the intersection of all states $S \in \overline{S_1}$. In computing this
intersection, we can associate with each new state in the object protocol
representing the intersection to those states in the original protocol it
was derived from. Wherever one of these associated sets contains a state
$S \in \overline{S_2}$, all of the states in that set must be in 
$\overline{S_2}$. Finally, all end states $S \in \overline{S_2}$ must
exist in one of the associated sets.

An example of this is shown in \figref{fig:sm1} for the protocol $O_1$.
If the input of the effect type is $O_1@\{B,D\}$ then the only valid
outputs are either $O_1@\{B,D\}$ (no calls are made),
$O_1@\{B,C\}$ (exactly one call to $b$ is made), or
$O_1@\{B,C,D\}$ (the union of those two options). The effect type
$O_1@\{B,D\} \gg O_1@\{B\}$ would be correctly identified as invalid.

\subsection{Interpreting effect types with subtyping}

The typing judgement for a function call permits a passed parameter to
be a subtype of the input portion of the associated effect type.
Determining the type of the parameter after after the
function call, in particular where the parameter is an object, 
requires some thought to avoid losing type information unnecessarily.

Consider a function call $f(x)$ where \\
$f : (O_2@A \gg O_2@\{E,H\}) \rightarrow \top$ and
$x : O_1@A$, where $O_1$ and $O_2$
are the state machines from \figref{fig:principalex}.
It is clear that $O_1@A <: O_2@A$, therefore by definition any possible
sequence of method calls that $f$ could call on $x$ would be valid.
What should be the type of $x$ after the call, given the information that
the effect type gives?

A very naive but safe choice would be to choose $O_2@\{E,H\}$. However,
this will often result in an unacceptable loss of type information,
reducing the number of valid programs which can be typed within the system.
Based on the effect type,
we know that the function will interact with the $x$ using
one of the possible sequences of methods calls from $O_2@A$ to 
$O_2@E$ or $O_2@H$. There must exist some
$\overline{S_3}$ such that $\ot{O_1}{S_3} \precsim O_2@\{E,H\}$.
One possible solution would be to choose
$\overline{S_3} = \{ S | O@S \precsim \ot{O_2}{S_4} \}$. 
This is still liable to lose type information however --- given the object 
protocols from \figref{fig:principalex}, this would produce
$\overline{S_3} = \{ A, B, C, D \}$, as there are no possible method calls
on states $E$ or $H$ of $O_2$ (all object types trivially simulate the
empty object type $\{ S\{\} \}@S$). This chosen solution also ignores the
path taken to $O_2@E$ or $O_2@H$, and so may contain states
that are actually unreachable from $O_1@A$.

It is clear that our choice of $S_3$ must relate the set of possible method
call sequences between $O_2@A$ and $O_2@\{E,H\}$. For this we
define a relation between pairs of object types
$(O_1@\overline{S}, O_2@\overline{S_2} \lesssim 
(O_1@\overline{S_3}, O_2@\overline{S_4})$, which we call the
{\it path relation}, which defines equivalent paths through the
respective object protocols. The definition of the path relation
is given in \figref{fig:simreach}.

@@

\subsection{Principal typings \& types}

A typing of the term $t$ is the triple $\tau = (\Gamma,T,\Gamma')$ such that
$\typerule{\Gamma}{t}{T}{\Gamma'}$. Typings can be ordered using the subtyping
relation. Given two typings $\tau_1 = (\Gamma_1,T_1,\Gamma_1')$ and
$\tau_2 = (\Gamma_2,T_2,\Gamma_2')$ of $t$ we say that $\tau_2$ is 
{\it stronger} than $\tau_1$, written $\tau_1 \leq \tau_2$, if:
